Learning an Intervention via Causal Inference

### 2016-01-31 causal-bandit/icml-2015/general.tex
starting to clean up general section

We now consider single variable interventions on general causal graphs over variables $\boldsymbol{V}$. Let $Y \in \set{0,1}$ be the target variable and $\boldsymbol{X} = V/\set{Y \cup Desendants(Y)}$.

If all the variables in $\boldsymbol{X}$ are observable, we can compute the distribution $P(Y,\boldsymbol{X}|do(X_i = j)$ for any intervention $i,j$ via the truncated product formula \cite{Pearl2000} 3.11

\eq {
\P{\boldsymbol{x},y|do(X_i = j)} = 
\begin{cases}
\frac{\P{\boldsymbol{x},y}}{\P{x_i=j|\calP_i(\boldsymbol{x})}} & \text{if} x_i = j  \\
0 & \text{otherwise}
\end{cases}
}

Where $\calP_i : \set{0,1}^N \to \set{0,1}^{N_i}$ is the
projection that selections the $N_i$ parents of variable $i$.

@@

\begin{figure}[h]
\centering
\caption{Causal model for the confounded parallel bandits problem}.\label{fig:causalStructure_confounded}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}, hidden/.style={empty}]

 %nodes

\node[main node](1){$X_{1}$};
\node[main node, right=of 1](2){$X_{2}$};
\node[hidden, right=of 2](3){$...$};
\node[main node, right=of 3](4){$X_{N}$};
\node[main node, below right=of 2](5){Y};
\node[main node, above=of 3](6){$Z$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (5)
    	(2) edge (5)
    (4) edge (5)
    (6) edge (1) edge (2) edge(4);

\end{tikzpicture}
\end{figure}

>>>>

We now consider single variable interventions on general causal graphs over binary variables $Y, X_1,\ldots,X_N$. 
As before $Y$ is the target variable and $\P{Y=1|X} = r(X)$ is the expected reward given $X$.
We let $\calP_i \subseteq \set{1,\ldots,N}$ denote the parents of $i$ so the joint distribution over $X$ may be written as
\eq{
\P{X = x} = \prod_{i=1}^N \P{X_i = x_i|X_j = x_j \text{ for all } j \in \calP_i}\,.
}
Throughout this section we are going to assume that the causal structure and conditional distributions of $X$ are known,
which allows us to compute the above quantity for arbitrary $x$. As in the previous sections we allow interventions on
a single variable by setting $X_i = j$ for any $i \in \set{1,\ldots,N}$ and $j \in \set{0,1}$. Then for $x \in \set{0,1}^N$ 
with $x_i = j$ we have
\eq{
&\P{X = x|do(X_i = j)} \\
&\qquad =\prod_{k \neq i} \P{X_k = x_k|X_j = x_j \text{ for all } j \in \calP_k}\,.
}
As before the expected reward from intervening by setting variable $X_i = j$ is
\eq{
\mu_{ij} = \sum_{x \in \Omega} \P{X = x|do(X_i = j)} r(x)\,,
}
where $\Omega = \set{0,1}^N$. The problem of learning the optimal intervention is now significantly complicated
by the dependencies between the random variables. Learning from observational data is no longer straightforward
because variables can be correlated through their parents. Consider, for example, the causual model below where $X_1$
deterministically causes $X_2$ and $X_3$, which ensures that $X_2 = X_3$ for all observational data. Now suppose $r(X) = \ind{X_2 \neq X_3}$,
then intervening on either $X_2$ or $X_3$ is optimal, but this positive is never observed from observational samples only. This is despite
the fact that $\P{X_2 = 1} = \P{X_3 = 1} = \frac{1}{2}$.
\begin{figure}[h]
\centering
\caption{Causal model for the confounded parallel bandits problem}.\label{fig:causalStructure_confounded}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}, hidden/.style={empty}]
\node[main node](1){$X_1$};
\node[main node, below left=of 1](2){$X_2$};
\node[main node, below right=of 1](3){$X_3$};
\node[main node, below right=of 2](4){Y};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2)
    (1) edge (3)
    (3) edge (4)
    (2) edge (4);
\end{tikzpicture}
\eq{
\P{X_1 = 1} &= \frac{1}{2} \\
\P{X_2 = j|X_1 = j} &= 
\P{X_3 = j|X_1 = j} = 1\,.
}
\end{figure}

@@

### 2016-01-31 causal-bandit/icml-2015/setup.tex
starting to clean up general section

A learner for a casual bandit problem is given the casual model's graph $\mathcal{G}$, a set of \emph{observable} variables $\mathcal{O} \subseteq \mathcal{X}$, a set of \emph{allowed interventions} $\mathcal{A}$, a \emph{reward variable} $Y$. 
The following game is then played over $T$ rounds.
In round $t$, the learner selects an allowed intervention $do(\vec{X}_t = \vec{x}_t)$ from $\mathcal{A}$.
Values for all non-intervened variables $\vec{X}^c_t$ are sampled from $\P{\vec{X}^c_t | do(\vec{X}_t = \vec{x}_t)}$. Finally, the learner is shown the sampled values of all the observable variables in $\vec{X}^c_t \cap \mathcal{O}$ and receives the reward $Y_t$.

@@

\todom{Mention pruning of actions?}
We note that classical $K$-armed stochastic bandit problem can be recovered in our framework by considering a simple causal model with one edge connecting a single variable $X$ that can take on $K$ values to a reward variable $Y = r(X) + \eta$ for some arbitrary but unknown, real-valued function $r$ and a noise term $\eta$ that may depend on $X$.
The set of allowed interventions in this case is $\mathcal{A} = \{ do(X = k) \colon k \in \{1, \ldots, K\}\}$.
Conversely, any causal bandit problem can reduced to a classical stochastic $|\mathcal{A}|$-armed bandit problem by treating each possible intervention as an arm and ignoring all sampled values for the observed variables except for the reward.
Intuitively though, one would expect to perform better by making use of the extra observations.

>>>>

A learner for a casual bandit problem is given the casual model's graph $\mathcal{G}$, a set of \emph{observable} variables $\mathcal{O} \subseteq \mathcal{X}$, a set of \emph{allowed interventions} $\mathcal{A}$, a \emph{reward variable} $Y$. 
The following game is then played over $T$ rounds.
In round $t$, the learner selects an allowed intervention $do(\vec{X}_t = \vec{x}_t)$ from $\mathcal{A}$.
Values for all non-intervened variables $\vec{X}^c_t$ are sampled from $\P{\vec{X}^c_t | do(\vec{X}_t = \vec{x}_t)}$. Finally, the learner is shown the sampled values of all the observable variables in $\vec{X}^c_t \cap \mathcal{O}$ and receives the reward $Y_t \in \set{0,1}$.

@@

\todom{Mention pruning of actions?}
\todot{I don't understand this at all. Later we are assuming $X_1,\ldots,X_N$ are binary and there are $N$ of them}
We note that classical $K$-armed stochastic bandit problem can be recovered in our framework by considering a simple causal model with one edge connecting a single variable $X$ that can take on $K$ values to a reward variable $Y \in \set{0,1}$ where $\P{Y = 1|X} = r(X)$ for some arbitrary but unknown, real-valued function $r$.
The set of allowed interventions in this case is $\mathcal{A} = \{ do(X = k) \colon k \in \{1, \ldots, K\}\}$.
Conversely, any causal bandit problem can reduced to a classical stochastic $|\mathcal{A}|$-armed bandit problem by treating each possible intervention as an arm and ignoring all sampled values for the observed variables except for the reward.
Intuitively though, one would expect to perform better by making use of the extra observations.
